Time to stop; agent-based modelling of chemoaffinity with competition

### 2022-08-11 paper/main.tex
Current state of paper from Overleaf

% Use the article doc class, with an 11 pt basic font size
%\documentclass[11pt, a4paper, draft]{article}
\documentclass[11pt, a4paper]{article}
% Nice citations
\usepackage{natbib}
% Set margins
\usepackage[margin=2.5cm]{geometry}
% Multilingual support
\usepackage[english]{babel}
% Nice mathematics and Left-right harpoons for kinetic equations
\usepackage{amsmath,mathtools}
% Control over maketitle
\usepackage{titling,titlesec}
% Ability to use colour in text
\usepackage[usenames]{color}
\definecolor{grey}{rgb}{0.6, 0.6, 0.63}
\definecolor{black}{rgb}{0, 0, 0}
% For the \degree symbol
\usepackage{textcomp,gensymb}
% Allow includegraphics and wrapped figures
\usepackage{graphicx,wrapfig,subfigure}
\usepackage[outercaption]{sidecap}
\usepackage{csquotes}

@@

\emph{We show that a true chemoaffinity model based on gradient following (via receptor-ligand interactions) combined with receptor-ligand-mediated axonal repulsion, in which information is only transferred by nearest-neighbour interactions, can account for a broad range of experimental results.}

@@

Sperry's chemoaffinity model is an influential theory of how axons from neurons that originate in one tissue can find their way to specific locations in a target tissue during early brain development \citep{sperry_chemoaffinity_1963}.
%
The central proposal of the theory is that growing axons find their location in a destination tissue by means of `labels of a cytochemical nature'.
The chemical affinity between labels in source and destination tissues allows axons to distinguish correct target locations from incorrect ones.
%
This key idea does not, on its own, explain how an axon can find the correct \emph{direction} in which it should move to reach its target.
An additional component of Sperry's theory \citep{gaze?} suggested that the axon could determine this direction by evaluating local information in gradients of gene expression spanning the target tissue, giving rise to chemotaxis of the axonal growth cones.
%
Sperry's theory was given robust support by the discovery of the ephrin ligands and their receptors \citep{cheng_complementary_1995,drescher_vitro_1995} which have been shown to form into graded expression fields in the retina \citep{braisted_graded_1997}, tectum \citep{braisted_graded_1997,feldheim_genetic_2000} as well as in other sensory systems, such as the somatosensory system \citep{vanderhaeghen_mapping_2000}.
%
The Ephrin ligands have a clear effect on axonal outgrowth, as shown in-vitro \citep{cheng_complementary_1995,drescher_vitro_1995,hansen_retinal_2004} and by in-vivo \citep{frisen_ephrin-a5_1998,rodger_transient_2000,mann_topographic_2002,hindges_ephb_2002} studies.

@@

However, while such models have \emph{assumed} the effects of chemotaxis, the key assumption of the chemoaffinity model, that axons sample and evaluate positional information only \emph{locally}, is yet to be tested directly.
%
Consider, for example, an important model by Goodhill \& Simpson, which shows how [a certain set of assumptions about the form of the] competition between axons can account for patterns of innervation in the developing tectum that result from ablating/rotating sections of the retina \citep{simpson_simple_2011}.
%
The formulation of this model includes a term that at every time-step moves the simulated axons in the direction of a pre-specified target location.
Thus the model implicitly assumes a global supervisor, capable of evaluating the current state of each axon, computing a direction vector to correct its trajectory, and communicating that information to each axon.
%
This model was constructed primarily to establish the effects of competition, and to this end is elegantly formulated.
%
Nevertheless, the key assumptions of the chemoaffinity model, and specifically the proposal that axons find their way via evaluating only local information (chemotaxis), are yet to be investigated directly.

The aim of the current study is therefore to construct a model that represents the assumptions of the chemoaffinity model explicitly.

@@

Indeed, our simulation results show that the local mechanisms of axonal information-processing that are assumed by the chemoaffinity model provide a remarkably robust solution to the problem of topographic map development.

@@

In an agent-based model of chemotaxis and competition, the position, $\mathbf{x}_t$, of each branch $b$ on a simulated tectum can be updated according to $\mathbf{x}_{t+1} = \mathbf{x}_{t} + \mathbf{M}$ with the movement vector $\mathbf{M}$ given by
\begin{equation} \label{e:mv2}
 \mathbf{M} = \mathbf{G} +  \mathbf{X} + \mathbf{B}.
\end{equation}
$\mathbf{G}$ is the movement due to the chemotaxis effect and $\mathbf{X}$ is the movement due to competitive interactions with other branches.
$\mathbf{B}$ is a movement applied close to the tissue boundary that ensures all branches remain within the tissue.

@@

Each retinal ganglion cell (RGC) projects $M=4$ growth cones (referred to as \emph{branches}) which carry a set of $N$ receptors, $r_i$ indexed by $i$ and expressed at levels determined by the cell soma's location on the retinal grid.
($N$ ligands, $l_i$, are also expressed by RGCs, but these are not used for chemotaxis.)
The expression of each receptor varies only with respect to one spatial dimension.
%
We defined 4 receptor expression gradients arranged in orthogonal pairs with the gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, was orthogonal to $r_3$ (Fig.\,\ref{f:ex}B).
%
Because there is convincing evidence that EphA and EphB receptors are expressed in exponentially increasing patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004}, we use an exponential form for retinal receptor expressions, in common with other modelling studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
We adopt the same precise form for the retinal receptor expression as \citet{simpson_simple_2011}, $f(x,y) = 1.05 + 0.26 \exp(2.3 u)$, where $u$ gives the direction with which the expression increases and is substituted by $(1-x)$ for $r_0$, $(1-y)$ for $r_1$, $x$ for $r_2$ and by $y$ for $r_3$ (Fig.\,\ref{f:ex}B).

@@

 We modelled competition as distance-based competitive movement vector, $\mathbf{X}$, which is `switched on' via receptor-ligand signalling:
%
\begin{equation}
\mathbf{X} = \frac{m_{\!_X}}{|B_{b}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W\,Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}).
\end{equation}
%
$W$ is the distance based weighting ($W = 1-\frac{d_{kb}}{2r_{\!_X}}~\mathrm{if}~  d_{kb}\leq 2r_{\!_X}$) and $B_{b}$ is the set of branches that are within interaction distance of $b$. $m_{\!_X}$ is a scalar parameter, $nM$ is the total number of branches and $\hat{\mathbf{x}}_{kb}$ is the unit vector from $k$ to $b$.
%
$Q$ is a signalling threshold function which depends on the distance $d_{kb}$ between two branches $b$ and $k$, the receptor expression on $b$ ($\mathbf{r}_b$) and the ligand expression on $k$ ($\mathbf{l}_k$):
%
\begin{equation}
Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}) = \begin{cases}
                 0 & \mathrm{if}~-F_i\,r_{i,b}\,l_{i,k} <
                 s,\,\forall{i}~\mathrm{or}~d_{kb} > 2r_{\!_X} \\
                 1 & \mathrm{otherwise.}
     \end{cases}
\end{equation}
%
$r_{\!_X}$ is the receptor-ligand interaction radius, $l_{i,k}$ (the expression of ligand $i$ on branch $k$) is the $i^{\mathrm{th}}$ element of $\mathbf{l}_k$ and $r_{i,b}$ is the $i^{\mathrm{th}}$ element of $\mathbf{r}_b$.
$F_i$, defined in Eq.\,\ref{e:G}, is -1 if the $r_{i}\,l_{i}$ interaction signals repulsion and +1 if it signals attraction.
$s$ is a positive signal threshold for interaction.
With the constraint $s>0$ comes the assumption that attractive interactions will have no effect on the competitive movement vector---a receptor-ligand pair that signals attraction cannot cause $Q$ to take the value 1.
$Q$ is set to 1 only if at least one repulsive signal exceeds $s$ and the distance from branch $b$ to branch $k$ is smaller than $2 r_{\!_X}$.

@@

The model has four free parameters; $m_{\!_G}$, $m_{\!_X}$, $r_{\!_X}$ and $s$. We fixed $m_{\!_G}$ to the value $0.0012$ and used the optimization to find the competition parameters that would balance this level of chemotaxis.
To find the best combination of the remaining three parameters, we reasoned that evolution would optimize the arrangement of the wildtype pattern.
We performed a simulated annealing optimization to minimize an objective function for the wildtype pattern (Fig.\,\ref{f:GJ}E) at simulation time 1000.

We used a combination of pattern metrics to define the objective function. $\epsilon$ is the mean distance error of the axon centroids, given by:
%
\begin{equation}\label{e:eps}
\epsilon = \sqrt{\frac{\sum_k^n (\mathbf{x}_{k} - \mathbf{x}'_{k})^2}{n}},
\end{equation}
%
where $\mathbf{x}_{k}$ is the location of the centroid of the branches of axon $k$ and $\mathbf{x}'_{k}$ the target tectal location for axon $k$.
$n$ is the number of RGC axons in the system.
%
$\epsilon$ tends to 0 as the pattern approaches the target layout.
However, two patterns may have the same non-zero value for $\epsilon$, despite one being qualitatively more disordered than the other.
For this reason, we also use the number of crossings in the fish net plot of the axon centroids (see Fig.\,\ref{f:GJ}D) as a second metric, called $\eta$, which tends towards an expected target number of crossings, $\eta_t$ as the pattern approaches the target layout. The annealing objective function, $\xi$, combined these metrics with $\xi = \epsilon^2 (1+\frac{\eta}{100})$.
The parameters returned by the simulated annealing were
$m_{\!_X} = 0.2078$, $r_{\!_X} = 0.0759$ and $s = 4.815$.

Fig.\,\ref{f:GJ} shows the simulation results with these parameters after 1500 time steps.
Fig.\,\ref{f:GJ}A indicates the colour code used for retinal cell position; dorsal retinal cells are indicated by green, nasal cells by red. Lightness gives position along the relevant axis so that cells at the dorsal-nasal corner are yellow and those at the ventral-temporal corner are black.
Fig.\,\ref{f:GJ}E shows the expected layout of retinal ganglion cell axons on the tectum.
The axons retain the colour of their originating cell on the retina.
Nasal cell axons should be arranged caudally; dorsal axons laterally; axons should be evenly spaced and the pattern should span the entire tectum.
Fig.\,\ref{f:GJ}, panels B--D show `fish net' plots of the average location of the 4 branches of each axon (the axon's centroid) at three timepoints.
Fig.\,\ref{f:GJ}B indicates the randomised placement of axon branches at the rostral end of the tectum as specified by the initial conditions (Eq.\,\ref{e:ic}).
In these plots, lines are drawn between axons that are expected to arrange themselves adjacently to one another by the end of the simulation.
The number of line crossings in this plot gives an indication of the disorder in the arrangement.
Fig.\,\ref{f:GJ}G shows the positions of the individual branches (i.e.~growth cones) at $t=1500$ that contribute to the plot in Fig.\,\ref{f:GJ}D.
By this time, the topological order matches that shown in Fig.\,\ref{f:GJ}E.
Opposing gradients have sorted the axons along the two axes to form a grid of cells.
The arrangement spans the full tectum due to competition between the growth cones (\textbf{ref supp. figure of chemo-only model}).
Fig.\,\ref{f:GJ}F shows the two pattern metrics; $\epsilon$ (distance error) and $\eta$ (number of crossings).
For the wildtype pattern, $\eta_t=0$, as in Fig.\,\ref{f:GJ}E.
Fig.\,\ref{f:GJ}H shows selected axons (one from the centre and one from each corner of the retina) and their position history.

@@

After finding an optimum set of parameters for the wildtype arrangement, we tested our model with a number of simulations of surgical manipulations (for reviews, see \citet{udin_formation_1988} and \citet{goodhill_retinotectal_1999}).
Figs.~\ref{f:GJsurg}A and \ref{f:GJsurg}B show the results of tectal graft rotation manipulations in which a square patch of tectal tissue is cut out and rotated $90\degree$ or $180\degree$ \citep{chung_observations_1978}.
%
To simulate the surgery, we rotated an $8\times8$ square of the tectum shown in Fig\,\ref{f:ex}D, carrying the ligand expressions with the rotation (resulting in Fig.\,\ref{f:trot90}, left column).
After the manipulation, we re-computed the tectal ligand expression gradients. Fig.\,\ref{f:trot90} demonstrates that the most significant change in the gradient is around the border of the graft, due to discontinuities in the ligand expression.
The manipulated gradients, interacting with unchanged axon branches, lead to a qualitative rotation of axons that were destined for the graft area, indicated by a 90$\degree$ or 180$\degree$ `twist' of the axon centroid colours.
Long connecting lines in the fish net show that a number of axons from near the edges of the tectum have been drawn into the graft region.
The pattern metric, $\epsilon$, is shown in the right hand graphs of Fig.\,\ref{f:GJsurg}A \& B.
We plot $\epsilon$ (the distance error) for the entire map (blue) along with the error for the graft region (gold) and the surround region (red).
The overall distance error for the rotation manipulations (0.182 or 0.2752)  is greater than for the wildtype system (0.04).
The pattern within the graft exhibits greater average distance error than for the surrounding tissue.
The system does not achieve a perfect reproduction of the target layout suggested by the experimental result (which is shown in the inset) but it is in qualitative agreement.
%
Fig.\,\ref{f:GJsurg}C shows the result of a similar manipulation in which two grafts are cut out of the tectum and swapped. \textbf{Fixme: which of the two grafts does the RH graph refer to?}

>>>>

% Use the article doc class, with an 11 pt basic font size
%\documentclass[11pt, a4paper, draft]{article}
\documentclass[11pt, a4paper]{article}
% Nice citations
\usepackage{natbib}
% Set margins
\usepackage[margin=2.5cm]{geometry}
% Multilingual support
\usepackage[english]{babel}
% Nice mathematics and Left-right harpoons for kinetic equations
\usepackage{amsmath,mathtools}
% Control over maketitle
\usepackage{titling,titlesec}
% Ability to use colour in text
\usepackage[usenames]{color}
\definecolor{grey}{rgb}{0.6, 0.6, 0.63}
\definecolor{black}{rgb}{0, 0, 0}
% For the \degree symbol
\usepackage{textcomp,gensymb}
% Allow includegraphics and wrapped figures
\usepackage{graphicx,wrapfig,subfigure}
\usepackage[outercaption]{sidecap}
\usepackage{csquotes}
% For full page figures:
%\usepackage[CaptionAfterwards]{fitpage}

@@

\emph{We show that a true chemoaffinity model based on gradient-following (via receptor-ligand interactions) combined with receptor-ligand-mediated axonal repulsion, in which information is only transferred by nearest-neighbour interactions, can account for a broad range of experimental manipulations of the retinotectal system.}

@@

% Could use chemoaffinity as a coverall theory, comprising chemospecificity and chemotaxis.

Sperry's chemoaffinity model is an influential theory of how axons from neurons that originate in one tissue can find their way to specific locations in a target tissue during early brain development \citep{sperry_reestablishment_1942,sperry_visuomotor_1943,sperry_chemoaffinity_1963}.
%
The central proposal of the theory is that growing axons find their location in a destination tissue by means of `labels of a cytochemical nature'.
The chemical affinity between labels in source and destination tissues allows axons to distinguish correct target locations from incorrect ones.
%
This key idea (called chemospecificity) does not, on its own, explain how an axon can find the correct \emph{direction} in which it should move to reach its target.
An additional component of Sperry's theory \citep{stevens_handbook_1950,sperry_problems_1955,sperry_chemoaffinity_1963} suggested that the axon could determine this direction by evaluating local information in gradients of gene expression spanning the target tissue, giving rise to chemotaxis of the axonal growth cones.

Much of Sperry's work was carried out on the retinotectal system, which is ideal for the study of ordered axonal development in fish, amphibia and birds, because in these animals, patterning of axons across the tectum is largely complete prior to the onset of neural activity (\textbf{find citation}).
In this system a two-dimensional map of the retina is developed on the surface of the optic tectum such that the axons from adjacent retinal ganglion cells (RGCs) find their way to adjacent locations on the tectum.
The retinotectal map is amenable to a number of surgical manipulations especially in amphibia, in which re-growth of the projection can be observed after surgical changes to the sensory apparatus (such as ablation or rotation of the retina) or the map substrate (the tectum).
This allowed Sperry and other researchers to carry out a wide range of experimental manipulations, the results of which informed the chemoaffinity theory.

In the 1990s, the theory of chemoaffinity and chemotaxis was given robust support by the discovery of the ephrin ligands and their receptors \citep{cheng_complementary_1995,drescher_vitro_1995} which have been shown to form into graded expression fields in the retina \citep{braisted_graded_1997}, tectum \citep{braisted_graded_1997,feldheim_genetic_2000} as well as in other sensory systems, such as the somatosensory system \citep{vanderhaeghen_mapping_2000}. % And sounds?
%
The Ephrin ligands have a clear effect on axonal outgrowth, as shown in-vitro \citep{cheng_complementary_1995,drescher_vitro_1995,hansen_retinal_2004} and by in-vivo \citep{frisen_ephrin-a5_1998,rodger_transient_2000,mann_topographic_2002,hindges_ephb_2002} studies.
Two classes of ephrins have been identified, labelled `A' and `B'.
Ephrin A and its receptors mediate a repulsive interaction which contributes to an ordering of RGC axons on the tectum along the rostral-caudal axis. Ephrin B acts on medial-lateral ordering with an attractive interaction.
Together, the ephrins are able to provide a coordinate system on the tectum to guide growing axons, but precisely how this positional information is used by axonal growth cones is not yet fully understood.

Quantitative models exploring axonal ordering were already in existence by the time of the discovery of the ephrins.
The importance of chemical gradients to biological processes was already well known from the study of slime moulds \citep{bonner_evidence_1947}.
Sperry himself discussed chemical gradients in relation to the problem of neuronal pathfinding \citep{sperry_problems_1955}, albeit qualitatively.
Some time after Sperry's active period, but before the ephrins were identified, \citet{gierer_development_1981,gierer_model_1983,gierer_directional_1987} made one of the few quantitative investigations of a gradient-based model of the retinotectal projection. He presented gradient following as a special case of a general class of \emph{potential models}.
In a potential model, the interactions between the properties of a growing axon originating from a location $\mathbf{u}$ on the retina and its environment at location $\mathbf{x}$ on the tectum set up some potential function, $p(\mathbf{x},\mathbf{u})$.
The axon then grows or moves to minimize $p$, which causes it to find its target location. This model successfully creates the wildtype map because each axon originates with a different $\mathbf{u}$ and has its minimum $p(\mathbf{x},\mathbf{u})$ in the correct target location.
This model forms a mathematical expression of Sperry's theory. Because it is quantitative, it allowed Gierer to identify two features of the theory: i) A \emph{pair} of opposing gradients is needed to form a potential function in one dimension (and \emph{two} pairs to create a 2D potential function). ii) Tectal or retinal ablation would require $p$ to be re-specified to reproduce the experimental results that fibres from an ablated retina will generate a map across the entire tectum and fibres from an intact retina will form a compressed map on an ablated tectum \citep{attardi_preferential_1963,schmidt_retinal_1978,schmidt_expansion_1978}.
Gierer addressed the second problem by introducing a form of competition between fibres, postulating that $p$ would accumulate locally at a rate proportional to the fibre density in that location and he noted that this was equivalent to a more direct fibre-fibre competition.
He demonstrated, in a one-dimensional computer simulation, that this form of competition would lead to the expected response to ablations \citep{gierer_model_1983}.

Gierer's simulations confirmed the validity of this basic idea, which we will refer to as \emph{differential chemotaxis}, for the problem of map formation with respect to a single spatial dimension \citep{gierer_model_1983,karbowski_model_2004}.
These models have also helped establish how competition between axons for space in the target tissue can additionally account for many of the known effects of surgical manipulation on map development.

Simulations of differential chemotaxis and axonal competition have recently been extended to investigate map development in two spatial dimensions, revealing how these two ingredients can give rise to realistic patterns of synaptic innervation \citep{james_modelling_2020}.
These models are formulated as systems of partial differential equations (PDE), and solved through numerical integration on an underlying lattice structure representing the target tissue.
In general, this approach makes mathematical analysis of system dynamics tractable, and as such PDE modelling has played a central r\^ole in the development of theories of pattern formation (see \citet{krause_modern_2021} for recent examples).
% Not sure about this:
However as the number of terms involved in PDEs necessarily grows large in systems representing differential chemotaxis, these systems quickly become impossible to analyse, making this potential advantage of such modelling redundant.
Additionally, the numerical simulation of a PDE system can become intractable, because each axonal projection must be represented by several state variables, each of which require solution across the entire spatial domain (which may consist of many thousands of elements).
% in james_modelling we did 41 projections. I found that trying to build a retinotectal system with about 100 projections was already intractable.

% But know we can motivate agent based modelling because *simulation* of PDEs for differential chemotaxis is intractable.
An alternative simulation approach is to treat each of $n$ axons as an agent that interacts with its environment and with the other $n-1$ axons.
The computation of such a system (see for example, \citet{simpson_simple_2011}) is dominated by the $O(n^2)$ axon-axon interactions and so a limit will be approached as $n\rightarrow \infty$.
We found that agent based simulations of $O(10^3)$ interacting axons can be achieved on a standard desktop computer whereas an equivalent PDE system would fail with about 100 axon projections.

\emph{PDEs too complex (can't do maths analysis), so need agent based. Can represent more biological detail and ask how that affects the systems level.}

\emph{List the questions that emerge from the discussion above}

\emph{Can Gierer model explain ALL the manipulations that S and G did?}

\emph{Allows us to make a specific representation of Gierer's model in which the receptor-ligand mechanism is elaborated.}

\textbf{xxxxxxxxxxxxxxx}

@@

However, while such models have \emph{assumed} the effects of chemotaxis, the key assumption of the chemoaffinity model, that axons sample and evaluate positional information only \emph{locally}, has yet to be tested directly.
%
Consider, for example, an important model by Goodhill \& Simpson, which shows how [a certain set of assumptions about the form of the] competition between axons can account for patterns of innervation in the developing tectum that result from ablating/rotating sections of the retina \citep{simpson_simple_2011}.
%
The formulation of this model includes a term that at every time-step moves the simulated axons in the direction of a pre-specified target location.
Thus the model implicitly assumes a global supervisor, capable of evaluating the current state of each axon, computing a direction vector to correct its trajectory, and communicating that information to each axon.
%
This model was constructed primarily to establish the effects of competition, and to this end is elegantly formulated.
%
Nevertheless, the key assumptions of the chemoaffinity model, and specifically the proposal that axons find their way by evaluating only local information (chemotaxis), are yet to be investigated directly.

The aim of the current study is therefore to construct a model that represents the assumptions of the chemotaxis model explicitly and to investigate how this mechanism behaves when combined with competitive interactions.

@@

Indeed, our results show that the local mechanisms of axonal information-processing that are assumed by the chemoaffinity model provide a remarkably robust solution to the problem of topographic map development.

@@

In an agent-based model of chemotaxis and competition, the position, $\mathbf{x}_t$, of each branch $b$ on a simulated tectum can be updated according to $\mathbf{x}_{t+1} = \mathbf{x}_{t} + \mathbf{M}$ with the movement vector $\mathbf{M}$ given by
\begin{equation} \label{e:mv2}
 \mathbf{M} = \mathbf{G} +  \mathbf{X} + \mathbf{B}.
\end{equation}
$\mathbf{G}$ is the movement due to the chemotaxis effect and $\mathbf{X}$ is the movement due to competitive interactions with other branches.
$\mathbf{B}$ is a movement applied close to the tissue boundary that ensures all branches remain within the tissue \citep{holt_target_1998}.

@@

Each retinal ganglion cell (RGC) projects $M=4$ growth cones (referred to as \emph{branches}) which carry a set of $N$ receptors, $r_i$, indexed by $i$ and expressed at levels determined by the cell soma's location on the retinal grid.
($N$ ligands, $l_i$, are also expressed by RGCs, but these are not used for chemotaxis.)
The expression of each receptor varies only with respect to one spatial dimension.
%
We defined 4 receptor expression gradients arranged in orthogonal pairs with the gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to $r_0$, was orthogonal to $r_3$ (Fig.\,\ref{f:ex}B).
%
Because there is convincing evidence that EphA and EphB receptors are expressed in exponentially increasing patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004}, we use an exponential form for retinal receptor expressions, in common with other modelling studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
We adopt the same precise form for the retinal receptor expression as \citet{simpson_simple_2011}, $f(x,y) = 1.05 + 0.26 \exp(2.3 u)$, where $u$ gives the direction with which the expression increases and is substituted by $(1-x)$ for $r_0$, $(1-y)$ for $r_1$, $x$ for $r_2$ and by $y$ for $r_3$ (Fig.\,\ref{f:ex}B).

@@

 We modelled competition as a distance-based competitive movement vector, $\mathbf{X}$, which is a sum of interactions that are `switched on' via receptor-ligand signalling:
%
\begin{equation} \label{e:X}
\mathbf{X} = \frac{m_{\!_X}}{|B_{b}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W\,Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}).
\end{equation}
%
$W$ is the distance based weighting ($W = 1-\frac{d_{kb}}{2r_{\!_X}}~\mathrm{if}~  d_{kb}\leq 2r_{\!_X}$, otherwise $0$) and $B_{b}$ is the set of branches that are within interaction distance of $b$. $m_{\!_X}$ is a scalar parameter, $nM$ is the total number of branches and $\hat{\mathbf{x}}_{kb}$ is the unit vector from $k$ to $b$.
%
$Q$ is a signalling threshold function which depends on the distance $d_{kb}$ between two branches $b$ and $k$, the receptor expression on $b$ ($\mathbf{r}_b$) and the ligand expression on $k$ ($\mathbf{l}_k$):
%
\begin{equation}
Q(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}) = \begin{cases}
                 0 & \mathrm{if}~-F_i\,r_{i,b}\,l_{i,k} <
                 s,\,\forall{i}~\mathrm{or}~d_{kb} > 2r_{\!_X} \\
                 1 & \mathrm{otherwise.}
     \end{cases}
\end{equation}
%
$r_{\!_X}$ is the receptor-ligand interaction radius, $l_{i,k}$ (the expression of ligand $i$ on branch $k$) is the $i^{\mathrm{th}}$ element of $\mathbf{l}_k$ and $r_{i,b}$ is the $i^{\mathrm{th}}$ element of $\mathbf{r}_b$.
$F_i$, defined in Eq.\,\ref{e:G}, is -1 if the $r_{i}\,l_{i}$ interaction signals repulsion and +1 if it signals attraction.
$s$ is a positive signal threshold for interaction.
With the constraint $s>0$ comes the assumption that attractive interactions will have no effect on the competitive movement vector---a receptor-ligand pair that signals attraction cannot cause $Q$ to take the value 1.
$Q$ is set to 1 only if at least one repulsive signal exceeds $s$ and the distance from branch $b$ to branch $k$ is smaller than $2 r_{\!_X}$.

@@

The model has four free parameters; $m_{\!_G}$, $m_{\!_X}$, $r_{\!_X}$ and $s$. We fixed $m_{\!_G}$ to the value $0.0012$ and used a numerical optimization process to find the competition parameters that would balance this level of chemotaxis.
We performed a simulated annealing optimization to minimize an objective function for the wildtype pattern (Fig.\,\ref{f:GJ}E) at simulation time 1000, reasoning that evolution would optimize the arrangement of this phenotype.

We used a combination of pattern metrics to define the objective function. $\epsilon$ is the mean distance error of the axon centroids, given by:
%
\begin{equation}\label{e:eps}
\epsilon = \sqrt{\frac{\sum_k^n (\mathbf{x}_{k} - \mathbf{x}'_{k})^2}{n}},
\end{equation}
%
where $\mathbf{x}_{k}$ is the location of the centroid of the branches of axon $k$ and $\mathbf{x}'_{k}$ the target tectal location for axon $k$.
$n$ is the number of RGC axons in the system.
%
$\epsilon$ tends to 0 as the pattern approaches the target layout.
However, two patterns may have the same non-zero value for $\epsilon$, despite one being qualitatively more disordered than the other.
For this reason, we also used the number of crossings in the fish net plot of the axon centroids (see Fig.\,\ref{f:GJ}D) as a second metric, called $\eta$, which tends towards an expected target number of crossings, $\eta_t$ as the pattern approaches the target layout. The annealing objective function, $\xi$, combined these metrics: $\xi = \epsilon^2 (1+\frac{\eta}{100})$.
The parameters returned by the simulated annealing were
$m_{\!_X} = 0.2078$, $r_{\!_X} = 0.0759$ and $s = 4.815$.

Fig.\,\ref{f:GJ} shows the simulation results with these parameters after 1500 time steps.
Fig.\,\ref{f:GJ}A indicates the colour code used for retinal cell position; dorsal retinal cells are indicated by green, nasal cells by red. Lightness gives position along the relevant axis so that cells at the dorsal-nasal corner are yellow and those at the ventral-temporal corner are black.
Fig.\,\ref{f:GJ}E shows the expected layout of retinal ganglion cell axons on the tectum.
The axons retain the colour of their originating cell on the retina.
Nasal cell axons should be arranged caudally; dorsal axons laterally; axons should be evenly spaced and the pattern should span the entire tectum.
Fig.\,\ref{f:GJ}, panels B--D show `fish net' plots of the average location of the 4 branches of each axon (the axon's centroid) at three timepoints.
Fig.\,\ref{f:GJ}B indicates the randomised placement of axon branches at the rostral end of the tectum as specified by the initial conditions (Eq.\,\ref{e:ic}).
In these plots, lines are drawn between axons that are expected to arrange themselves adjacently to one another by the end of the simulation.
The number of line crossings in this plot gives an indication of the disorder in the arrangement.
Fig.\,\ref{f:GJ}G shows the positions of the individual branches (i.e.~growth cones) at $t=1500$ that contribute to the plot in Fig.\,\ref{f:GJ}D.
By this time, the topological order matches that shown in Fig.\,\ref{f:GJ}E.
Opposing gradients have sorted the axons along the two axes to form a grid of cells.
The arrangement spans the full tectum due to competition between the growth cones (compare with the chemotaxis-only model in Fig.\,\ref{f:G}).
Fig.\,\ref{f:GJ}F shows the two pattern metrics; $\epsilon$ (distance error) and $\eta$ (number of crossings).
For the wildtype pattern, $\eta_t=0$, as in Fig.\,\ref{f:GJ}E.
Fig.\,\ref{f:GJ}H shows selected axons (one from the centre and one from each corner of the retina) and their position history.

@@

After finding an optimum set of parameters for the wildtype arrangement, we tested our model with a number of simulations of surgical manipulations (for reviews, see \citet{udin_formation_1988} and \citet{goodhill_retinotectal_1999}).
Figs.~\ref{f:GJsurg}A and \ref{f:GJsurg}B show the results of tectal graft rotation manipulations in which a square patch of tectal tissue is cut out and rotated $90\degree$ or $180\degree$ \citep{chung_observations_1978}.
%
To simulate the surgery, we rotated an $8\times8$ square of the tectum shown in Fig\,\ref{f:ex}D, carrying the ligand expressions with the rotation (resulting in Fig.\,\ref{f:trot90}, left column).
After the manipulation, we re-computed the tectal ligand expression gradients. Fig.\,\ref{f:trot90} demonstrates that the most significant change in the gradient is around the border of the graft, due to discontinuities in the ligand expression.
The manipulated gradients, interacting with unchanged axon branches, lead to a qualitative rotation of axons that were destined for the graft area, indicated by a 90$\degree$ or 180$\degree$ `twist' of the axon centroid colours.
Long connecting lines in the fish net show that a number of axons from near the edges of the tectum have been drawn into the graft region.
The pattern metric, $\epsilon$, is shown in the right hand graphs of Fig.\,\ref{f:GJsurg}A \& B.
We plot $\epsilon$ (the distance error) for the entire map (blue) along with the error for the graft region (gold) and the surround region (red).
The overall distance error for the rotation manipulations (0.182 or 0.2752)  is greater than for the wildtype system (0.04).
The pattern within the graft exhibits greater average distance error than for the surrounding tissue.
The system does not achieve a perfect reproduction of the target layout suggested by the experimental result (which is shown in the inset) but it is in qualitative agreement.
%
Fig.\,\ref{f:GJsurg}C shows the result of a similar manipulation in which two grafts are cut out of the tectum and swapped.
There is considerable disorder in the simulated pattern of axon centroids but a line of red/yellow agents that would been located caudally has advanced rostrally and a number of green agents are more caudal than in the wildtype arrangement, consistent with experimental observations.
% Note: the graph shows the pattern metric for a region that covers BOTH grafts and for the region that is not the two grafts.

Fig.\,\ref{f:GJsurg}D shows the result of a retinal ablation, in which the temporal half of the retina is surgically removed and the remaining RGC axons are allowed to develop on the unmanipulated tectum.
According to a pure chemoaffinity theory, the surviving retinal axons should retain their chemical labels and they would be expected to find their usual locations on the tectum, resulting in a tectum which is only half populated with axons.
The experimental observation is that the axons arrange themselves across the whole tectum, retaining their topographic order (inset, \citet{attardi_preferential_1963,schmidt_expansion_1978}).
The fish-net plot of the simulation indicates that the competition in the system successfully `stretches out' the surviving axons to achieve a close representation of the experimentally observed result.
In this figure, the right-hand graph shows the simulated position of each axon centroid on the rostro-caudal tectal axis plotted against its position on the nasal-temporal retinal axis in red triangles, with the experimental positions plotted in blue squares.
%
Fig.\,\ref{f:GJsurg}E shows the result of a tectal ablation. Here the retina is left un-modified, but the caudal half of the tectum is surgically removed. Experimental observation \citep{yoon_reorganization_1971,sharma_reformation_1972} shows that RGC axons `crush up' on the remaining rostral section, retaining topographic order and creating a complete map on the available tectal tissue. Our model clearly replicates this result, indicated by both fish-net and tectal/retinal axes graphs, because the chemotaxis component of the model is based on balanced, relative interactions between opposing gradients, rather than absolute labels or expression measurements.
%
A related ablation experiment is the so-called `mismatch', in which the temporal half of the retina is ablated along with the caudal part of the tectum to which it normally maps, again challenging the pure chemoaffinity model \citep{horder_retention_1971}. Our model replicates this manipulation (Fig.\,\ref{f:GJsurg}F) for the same reason as in the simple tectal ablation of Fig.\,\ref{f:GJsurg}E; opposing gradients remain on the surviving tectum allowing RGC axons to find their relative ordering.
%
The final surgical manipulation which we investigated was the `compound eye' experiment, in which two nasal retina halves are fused together \citep{gaze_retino-tectal_1963,fawcett_retinotectal_1982}. Within the compound retina there exist matched pairs of retinal ganglion cells, one from each nasal half, whose receptor and ligand expressions cannot be distinguished. These form overlaid maps, each of which spans the entire tectum (Fig.\,\ref{f:GJsurg}G).
